Testing gene methylation enrichment analysis approaches using simulated data

### 2023-10-15 figs/GSE158422_simulate.Rmd
sim

seed=100

# select gene sets to alter
gset_mod <- gset_selection(genesetdatabase=gs_symbols,seed=seed,gene_catalog=gene_catalog)

# select normal samples only
normal <- mval[,(1:(ncol(mval)/2)*2)]

# divide sets between hyper and hypomethylated
nset <- floor(length(gset_mod)/2)
set.seed(seed) ; gtup <-sample(gset_mod,48)
set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
gup <- unname(unlist(gtup))
gdn <- unname(unlist(gtdn))

# divide samples between ctrl and case
set.seed(seed) ; ctrl <- sample(1:ncol(normal),18)
set.seed(seed) ; case <- sample(setdiff(1:37,ctrl),18)
mval_ctrl <- mval[,ctrl]
mval_case <- mval[,case]

# select probes

>>>>

seed=100
frac_genes=0.5
frac_probes=0.5
normal_mval <- mval[,(1:(ncol(mval)/2)*2)]

incorp_dm <- function(genesetdatabase,myann,gene_catalog,mval,seed,frac_genes,frac_probes) {
  # select gene sets to alter
  gset_mod <- gset_selection(genesetdatabase=genesetdatabase,seed=seed,gene_catalog=gene_catalog)
  # divide gene sets between hyper and hypomethylated
  nset <- floor(length(gset_mod)/2)
  set.seed(seed) ; gtup <-sample(gset_mod,nset)
  set.seed(seed) ; gtdn <- sample(setdiff(gset_mod,gtup),nset)
  gup <- unname(unlist(gtup))
  gdn <- unname(unlist(gtdn))
  # make probe-gene vector
  probe2gene <- strsplit(myann$UCSC_RefGene_Name,";")
  names(probe2gene) <- rownames(myann)
  probe2gene <- unlist(probe2gene)
  # select probes hypermethylated
  set.seed(seed) ; gup2 <- sample(gup,floor(length(gup)*frac_genes))
  pup <- names(probe2gene[which(probe2gene %in% gup2)])
  set.seed(seed) ; pup2 <- sample(pup,floor(length(pup)*frac_probes))
  # select probes hypomethylated
  set.seed(seed) ; gdn2 <- sample(gdn,floor(length(gdn)*frac_genes))
  pdn <- names(probe2gene[which(probe2gene %in% gdn2)])
  set.seed(seed) ; pdn2 <- sample(pdn,floor(length(pup)*frac_probes))
  # divide samples between ctrl and case
  groupsize=floor(ncol(mval)/2)
  ncols <- ncol(mval)
  set.seed(seed) ; ctrl <- sample(1:ncols,groupsize)
  set.seed(seed) ; case <- sample(setdiff(1:ncols,ctrl),groupsize)
  mval_ctrl <- mval[,ctrl]
  mval_case <- mval[,case]
  # incorporate altered signals - change by +1 or -1
  mval_case[rownames(mval_case) %in% pup2,] <-  mval_case[rownames(mval_case) %in% pup2,] +1
  mval_case[rownames(mval_case) %in% pdn2,] <-  mval_case[rownames(mval_case) %in% pdn2,] -1
  mval2 <- cbind(mval_ctrl,mval_case)
  result <- list("mval"=mval2,"probes up"=pup2,"probes down"=pdn2,
    "genes up"=gup2,"genes down"=gdn2,
    "genesets up"=gtup,"genesets down"=gtdn)
  return(result)
}

This is how to use the function

```{r,run1sim}

sim <- incorp_dm(genesetdatabase=gs_symbols,myann=myann,gene_catalog=gene_catalog,
  mval=normal_mval,seed=100,frac_genes=0.5,frac_probes=0.5)

mval2 <- sim$mval
ncols <- ncol(mval2)
groupsize <- ncols/2
ss <- data.frame(colnames(mval2))
colnames(ss) <- "sample"
ss$case <- c(rep(0,groupsize),rep(1,groupsize))

d <- model.matrix(~ ss$case)
fit.reduced <- lmFit(mval2,d)
fit.reduced <- eBayes(fit.reduced)

summary(decideTests(fit.reduced))
dm <- topTable(fit.reduced,coef=ncol(d), number = Inf)
head(dm)
